Designing modular software for reproducible analyses: a case study in introductory statistics

### 2016-09-07 jcgs-paper-2016/paper.Rmd
Comments from meeting
Notes from meeting with Dr. Hofmann

---
title: "Introductory statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
  `intRo` is a web-based application for performing basic data analysis and statistical routines. Leveraging the power of R and Shiny, `intRo` implements common statistical functions in an extensible modular structure, while including a point-and-click interface for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: Interactivity, Programming Paradigms, Reactive Programming, Reproducibility, Statistical Software
bibliography: refs.bib
blind: false
---

# Introduction

The widespread adoption of R [@r-stat] as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student [@5359977]. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even if the instructor himself/herself uses R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms.

@@

Multiple software packages have recently been written in an attempt to spur interest in  R programming and statistics. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have. R Commander [@fox2005] and Deducer [@fellows2012] provide a graphical front-end to many statistical functions in R, but are not web-based and thus require local installation and configuration. iNZight Lite [@inzight] also attempts to expose students to data analysis without requiring programming knowledge, but does not include reproducible R code, and therefore has less of a focus on spurring interest in coding for students.

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on the design decisions of `intRo`, in particular, its modular and extensible structure. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

We've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo and hyperlinked within the application itself. The documentation describes the specific functionality provided by the default `intRo` modules.

Statistics courses with a lab component can most benefit from the use of `intRo`. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the student through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the student can easily navigate the `intRo` interface to obtain the results.

>>>>

---
title: "Designing modular software for reproducible analyses: a case study in introductory statistics"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
   Modular programming is a development paradigm that emphasizes self-contained, flexible, and independent pieces of functionality. However, the rise of web-based software applications has presented new challenges for designing an extensible, modular software system. In this paper, we outline a framework for designing such a system, with a focus on reproducibility of the results. We present as a case study a Shiny-based web application called `intRo`, that allows the user to perform basic data analyses and statistical routines. Finally, we highlight some challenges we encountered when combining modular programming concepts with reactive programming as used by Shiny.
keywords: Interactivity, Programming Paradigms, Reactive Programming, Reproducibility, Statistical Software
bibliography: refs.bib
blind: false
---

# Background

Modularity is a pervasive concept in computer science, extending from the design of systems
[@parnas1972criteria], to the design of software [@szyperski1996independently]. Modularity offers several advantages to both a developer and a user. In particular, functionality can be dynamically loaded and unloaded depending on the particular use case. Open source modular software opens up the possibility of extensions contributed by a wide array of programmers, which can allow the software to morph into areas that weren't anticipated early in development. In the statistics realm, R [@r-stat] is a prime example of the virtues of modular programming. As of this writing, The Comprehensive R Archive Network (CRAN) contains over 9000 source packages which can be installed and dynamically loaded in a particular session as needed.

Other statistics software also makes use of a number of these ideas. Microsoft Excel and JMP both include support for extensions, called macros and add-ins respectively, which allow programming routines to be written extending the base functionality of these programs. Compared with R, however, these programs don't maintain a large central repository of publicly available extensions on the level of CRAN. There are also software packages building upon R, and thus gaining the advantages of CRAN natively, such as R Commander [@fox2005] and Deducer [@fellows2012], which each provide a graphical front-end to many statistical functions in R. One thing these software packages all have in common is the requirement of local installation and configuration, which means certain operating systems and platforms will not support their use.

Modern web technologies have enabled a new generation of software packages that reside solely on the web, which eliminates the issue of local installation. Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. Several recent software packages have built up Shiny to provide a web-based system based on R. One such package is iNZight Lite [@inzight] which attempts to expose students to data analysis without requiring programming knowledge. Like most web-based systems, this does not include reproducible R code, which limits its useful in a scientific or academic setting. Another package is called Radiant [@radiant]

XXX Say something about reproducibility here - it has to be a goal to make an analysis reproducibility (baggerly cancer trials, forensic bioinformatics)

In our view, a modern software system should be **modular**, **extensible**, **web-based**, and foster **reproducibility**. We have developed a case-study application called intRo which we will use to illustrate our method of developing a system meeting these criteria. The paper is structured as follows: Section \ref{case-study-intro} describes the motivation behind intRo as an extensible system. section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on how we build intRo, and how it can be used to develop other modular systems. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of both intRo, and modular systems in general.

# Case Study: intRo

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student [@5359977]. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even if the instructor himself/herself uses R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms.

@@

Multiple software packages have recently been written in an attempt to spur interest in  R programming and statistics. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics.

These limitations led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text.

@@
